Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Full intro draft

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
The classic CDI case typically occurs soon after antibiotic use, which perturbs the
protective gut microbiota and allows _C. diff_ to proliferate [@kelly_can_2012].
Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
have also been shown to increase CDI susceptibilty and inhibit clearance  
[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
Diarrhea is the primary symptom, with some patients developing colitis, 
toxic megacolon, or requiring intensive care with an in-hospital mortality rate
of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
cases are associated with increased morbidity and mortality risk
[@napolitano_clostridium_2017;@kelly_can_2012].
CDI remains a significant burden on the US health care system with approximately
500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].

Many factors influence CDI severity, 
however, there is not a consensus on how to
define severity [@ressler_defining_2021]. 
disease-related complications definition [@mcdonald_recommendations_2007].
tbh gut microbiome may only play a small role.
Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.
Contribution of the gut microbiome.

prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
so how about predicting severity of infections to guide treatment.
also models on EHR to predict adverse outcomes (Li, Rao ribotype).
serum-based biomarker model (Dieterle mbio 2020).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

We chose to explore several different ways to define CDI severity (@fig-flowchart).
The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
however, it is easy to collect.

new dataset.

Two goals: investigate whether we can predict CDI severity based on OTU data to
inform how the gut microbiome may modulate severity (ML-based science: good
performance implies something about underlying biology), and determine whether
there is potential clinical value in OTU-based models.

@@

@@

Compare to EHR-based models.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be.
further work needed to characterize the costs and benefits.

@@

@@

We explore four different ways to define CDI cases as severe or not.
The IDSA definition of severe CDI is based on lab values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
The remaining definitions focus on the occurrence of adverse outcomes, which
may be more clinically relevant.
The all-cause severity definition defines a case as severe if ICU admission,
colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
cause of the adverse event.
The attributable severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
Finally, we defined a pragmatic severity definition that makes use of the
attributable definition when available and falls back to the all-cause definition
when chart review has not been completed, allowing us to use as many samples as
we have available while taking physicians' expert opinions into account where
possible.

>>>>

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
The classic CDI case typically occurs soon after antibiotic use, which perturbs the
protective gut microbiota and allows _C. difficile_ to proliferate [@kelly_can_2012].
Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
have also been shown to increase CDI susceptibilty and inhibit clearance
[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
Diarrhea is the primary symptom, with some patients developing colitis, 
toxic megacolon, or requiring intensive care with an in-hospital mortality rate
of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
cases are associated with increased morbidity and mortality risk
[@napolitano_clostridium_2017;@kelly_can_2012].
Patient risk factors for CDI-related morbidity and mortality include age > 65,
history of recurrent CDI, and co-morbid chronic illnesses [@ressler_defining_2021].
CDI remains a significant burden on the US health care system with approximately
500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].

There is a need for robust, accurate methods to identify patients at risk of
severe CDI outcomes.
When paired with treatment options that may reduce risk of severity, prediction
models can guide clinician decision-making to improve patient outcomes while
minimizing harms from unnecessary treatment.
Numerous scoring systems for predicting severe CDI outcomes based on patient
clinical factors have been developed, but none have validated well to external
datasets nor are any in use in routine clinical practice
[@chakra_prediction_2012;@perry_external_2022].
Machine learning (ML) is a promising approach that allows for use of thousands
of features to  classify samples and predict outcomes, rather than relying on
limited sets of human-curated variables.
Indeed, ML models trained on entire Electronic Health Record (EHR) data
have demonstrated improved performance over curated models
[@rao_clostridium_2015;@li_using_2019;@dieterle_systemic_2020].

Aside from patient factors encoded in EHRs,
the state of the patient gut microbiome is a promising factor to predict
severity, as the host microbiota can play either a protective or harmful
role in _C. difficile_ colonization, infection, and clearance.
Mouse studies have found that the initial state of the gut microbiome predicts
differences in clearance, moribundity, and cecal tissue damage in mice infected
with CDI [@tomkovich_initial_2020;@lesniak_gut_2022].
Identifying features of the human gut microbiota that promote or prevent severe
infections can guide further experiments to elucidate microbial mechanisms of
CDI severity, and incorporating these features into CDI severity models may
improve model performance to help guide clinical treatment decisions.

We set out to investigate whether ML models trained on features of the gut
microbiota can predict CDI severity in a human cohort,
whether the severity definition employed affects model performance,
and whether there is potential clinical value in deploying OTU-based models.
Stool samples from `r n_cases_first` CDI patients were collected on the day of
diagnosis and 16S rRNA gene amplicon sequencing was performed, followed by
clustering sequences into Operational Taxonomic Units (OTUs).
We then trained ML models to predict each of four severity definitions from
OTU relative abundances,
identified which microbial features contributed most to model performance, and 
conducted a proof-of-concept analysis of the potential clinical value of these
OTU-based models and compared these to prior EHR-based models.

@@

## CDI severity

There is not currently a consensus definition of CDI severity.
Some scoring systems such as the IDSA severity score focus on laboratory data
available during the course of CDI, while others focus on adverse outcomes of
CDI at 30 days after diagnosis [@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The IDSA definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
The remaining definitions focus on the occurrence of adverse outcomes, which
may be more clinically relevant.
The all-cause severity definition defines a case as severe if ICU admission,
colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
cause of the adverse event.
The attributable severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
Finally, we defined a pragmatic severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity for each definition and
determined how well OTU-based models perform for each of them.

@@

Compare to EHR-based models.
tbh gut microbiome may only play a small role.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
further work needed to characterize the costs and benefits.

@@

Two goals: investigate whether we can predict CDI severity based on OTU data to
inform how the gut microbiome may modulate severity (ML-based science: good
performance implies something about underlying biology), and determine whether
there is potential clinical value in OTU-based models.
new dataset.

@@

We chose to explore four different ways to define CDI cases as severe or not (@fig-flowchart).

- **IDSA**: A case is severe if serum creatinine level is greater than or equal
  to $1.5 mg/dL$ and the white blood cell count is greater than or equal to
  $15 k/\mu L$ on the day of diagnosis [@mcdonald_clinical_2018].
- **All-cause**: A case is severe if ICU admission, colectomy, or death
  occurred within 30 days of CDI diagnosis, regardless of the cause of the
  adverse event.
- **Attributable**: A case is severe if an adverse event of ICU admission,
  colectomy, or death occurred within 30 days of CDI diagnosis, and the adverse
  event was determined to be attributable to the CDI by two physicians who
  reviewed the medical chart [@mcdonald_recommendations_2007].
- **Pragmatic**: A case's severity is determined by the attributable definition
  if it is available, otherwise it is determiend by the all-cause definition.
